---
title: "Getting Started with mpindex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mpindex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is the MPI?

The **Multidimensional Poverty Index (MPI)** measures poverty not just by income, but by whether people are deprived across multiple areas of their lives — health, education, and living standards. A person or household is considered poor only if they fall short on *enough* of these areas at the same time.

The approach used here is the **Alkire-Foster (AF) method**, developed by Sabina Alkire and James Foster at the Oxford Poverty and Human Development Initiative (OPHI). It is the basis of the [Global MPI](https://ophi.org.uk/multidimensional-poverty-index/) published annually by OPHI and the UNDP.

The `mpindex` package makes it straightforward to compute the MPI from survey data in R.

---

## Installation

Install the released version from CRAN:

```{r, eval = FALSE}
install.packages("mpindex")
```

Or install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("yng-me/mpindex")
```

Then load the package:

```{r setup}
library(mpindex)
```

---

## The workflow at a glance

Computing the MPI with `mpindex` follows three steps:

1. **Define your indicators** — load a specification file that describes the dimensions, indicators, and weights.
2. **Specify deprivation cutoffs** — tell the package when a household is considered deprived on each indicator.
3. **Compute and explore the results** — run `compute_mpi()` to get the index, headcount ratio, intensity, and indicator contributions.

The sections below walk through each step using a built-in dataset.

---

## Step 1: Define your MPI specifications

### What goes in a specification file?

The specification file tells `mpindex` how your indicators are organized. It must contain these columns (column names are not case-sensitive):

| Column        | What it means |
|---------------|---------------|
| `Dimension`   | The broad domain (e.g. Health, Education) |
| `Indicator`   | The specific measure within that dimension |
| `Variable`    | The column name in your dataset that holds this measure |
| `Weight`      | How much this indicator contributes to the overall score |
| `Description` | *(optional)* A plain-language description of the indicator |

The package accepts `.csv`, `.xlsx`, `.json`, and `.txt` (tab-separated) files.

### Using the built-in Global MPI specification

`mpindex` ships with the Global MPI specification as an example. The `global_mpi_specs()` (formerly `global_mpi_specs()` which is now deprecated) shortcut loads it in one line:

```{r}
mpi_specs <- global_mpi_specs(uid = "uuid", unit_of_analysis = 'households')
```

The `uid` argument names the column in your dataset that uniquely identifies each household (the unit of analysis). In the built-in dataset, it is `"uuid"` (yours might be different).

Behind the scenes this is equivalent to:

```{r, eval = FALSE}
specs_file <- system.file("extdata", "global-mpi-specs.csv", package = "mpindex")
mpi_specs  <- define_mpi_specs(specs_file, uid = "uuid", unit_of_analysis = 'households')
```

The Global MPI covers 10 indicators across three dimensions, each weighted equally within its dimension:

```{r, echo = FALSE}
specs_file <- system.file("extdata", "global-mpi-specs.csv", package = "mpindex")
read.csv(specs_file) |>
  gt::gt() |>
  gt::tab_header(
    title    = "Global MPI — Dimensions, Indicators, and Weights",
    subtitle = "Source: OPHI MPI Methodological Note 49 (2020)"
  ) |>
  gt::tab_options(table.width = "100%", table.font.size = 12) |>
  gt::fmt_number(columns = 4, decimals = 3)
```

### Using your own specification file

If you are computing a custom MPI, with different sets of dimensions and indicators, create a CSV (or Excel) file that follows the same column structure and load it the same way:

```{r, eval = FALSE}
mpi_specs <- define_mpi_specs(
  mpi_specs_file = "path/to/my-specs.csv",
  uid = "household_id",
  poverty_cutoffs = 1/3 # default: a household needs >= 33% weighted deprivation score to be MPI poor
)
```

You can pass multiple cutoffs if you want to compare results across different poverty thresholds:

```{r, eval = FALSE}
mpi_specs <- define_mpi_specs(
  "path/to/my-specs.csv",
  uid = "household_id",
  poverty_cutoffs = c(0.20, 1/3, 0.50)   # 20%, 33%, 50%
)
```

---

## Step 2: Prepare your data

### The two built-in datasets

The package includes synthetic household survey data modelled on the Global MPI:

- **`df_household`** — one row per household, containing household-level variables.
- **`df_household_roster`** — one row per household member, containing individual-level variables (e.g. nutrition, school attendance).

```{r, message = FALSE, warning = FALSE}
library(dplyr)

glimpse(df_household)
```

```{r}
glimpse(df_household_roster)
```

> **Why two tables?** Some indicators (e.g. whether *any* child in the household is not attending school) come from individual-level data and must be collapsed to the household level. `mpindex` handles this automatically — you just tell it which table to use and how to collapse it.

---

## Step 3: Compute the MPI

### The main function: `compute_mpi()`

`compute_mpi()` takes your household data, your specifications, and a list of **deprivation cutoffs** — one per indicator. Each cutoff is a logical expression wrapped in `deprived()` that evaluates to `TRUE` when a household is deprived.

```{r}
mpi_result <- compute_mpi(
  df_household,
  mpi_specs = mpi_specs,
  deprivations = list(

    # --- Health ---
    nutrition = deprived(
      undernourished == 1 & age < 70,   # deprived if any member under 70 is undernourished
      .data        = df_household_roster,
      collapse_fn = max                 # household is deprived if any member is deprived
    ),
    child_mortality = deprived(with_child_died == 1),

    # --- Education ---
    year_schooling = deprived(
      completed_6yrs_schooling == 2,
      .data        = df_household_roster,
      collapse_fn = max
    ),
    school_attendance = deprived(
      attending_school == 2 & age %in% 5:24,
      .data        = df_household_roster,
      collapse_fn = max
    ),

    # --- Living Standards ---
    cooking_fuel   = deprived(cooking_fuel %in% c(4:6, 9)),
    sanitation     = deprived(toilet > 1),
    drinking_water = deprived(drinking_water == 2),
    electricity    = deprived(electricity == 2),
    housing        = deprived(
      roof %in% c(5, 7, 9) | walls %in% c(5, 8, 9, 99) == 2 | floor %in% c(5, 6, 9)
    ),
    assets = deprived(!(
      (asset_tv + asset_telephone + asset_mobile_phone + asset_computer +
         asset_animal_cart + asset_bicycle + asset_motorcycle +
         asset_refrigerator) > 1 &
        (asset_car + asset_truck) > 0
    ))
  )
)
```

The result is a named list:

```{r}
names(mpi_result)
```

### Disaggregating by a subgroup

Pass `by` to break results down by a grouping variable (e.g. urban vs. rural):

```{r, eval = FALSE}
compute_mpi(
  df_household,
  mpi_specs = mpi_specs,
  deprivations = list(...),
  by = class        # column in df_household
)
```

### The step-by-step alternative

If you prefer to build and inspect each deprivation indicator before combining them, use `define_deprivation()` to create each one separately, then pass them all to `compute_mpi_from_profile()`:

```{r, eval = FALSE}
dp <- list()

dp$nutrition <- df_household_roster |>
  define_deprivation(
    indicator = nutrition,
    cutoff = undernourished == 1 & age < 70,
    collapse_fn = max,
    mpi_specs = mpi_specs
  )

dp$drinking_water <- df_household |>
  define_deprivation(
    indicator = drinking_water,
    cutoff = drinking_water == 2,
    mpi_specs = mpi_specs
  )

# ... define all remaining indicators, then:
mpi_result <- compute_mpi_from_profile(df_household, dp, mpi_specs = mpi_specs)
```

This is useful when your data preparation for some indicators is complex and you want to examine intermediate results.

---

## Step 4: Explore the results

### Understanding the output

`mpi_result` is a named list with four components. Each component is itself a named list keyed by the poverty cutoff used — e.g. `k_33` for the 33% cutoff.

| Component | What it contains |
|-----------|-----------------|
| `$index` | The headline MPI value, headcount ratio (H), intensity (A), and sample size (n) |
| `$contribution` | Each indicator's percentage contribution to the overall MPI |
| `$headcount_ratio` | The share of households deprived on each indicator (uncensored and censored) |
| `$deprivation_matrix` | Row-level deprivation scores and indicator flags for every household |

### The headline MPI

The three key numbers are:

- **H (headcount ratio)** — what share of the population is multidimensionally poor.
- **A (intensity)** — among the poor, what fraction of the weighted indicators are they deprived in on average.
- **MPI** — the product H × A, capturing both *how many* people are poor and *how poor* they are.

```{r, eval = FALSE}
mpi_result$index$k_33
```

```{r, echo = FALSE}
mpi_result$index$k_33 |>
  gt::gt() |>
  gt::tab_header(title = "MPI — 33% Poverty Cutoff") |>
  gt::fmt_number(columns = 2:4, decimals = 3) |>
  gt::tab_options(table.width = "100%", table.font.size = 12)
```

### Indicator contributions

How much does each indicator drive the overall MPI? The contribution table answers this — values sum to 100% across all indicators.

```{r, eval = FALSE}
mpi_result$contribution$k_33
```

```{r, echo = FALSE}
gtx <- function(.gt, .decimals = 1, .offset = 0) {
  d01_cp <- 2:3  + .offset
  d02_cp <- 4:5  + .offset
  d03_cp <- 6:11 + .offset
  .gt |>
    gt::tab_spanner(label = "Health",           columns = d01_cp) |>
    gt::tab_spanner(label = "Education",        columns = d02_cp) |>
    gt::tab_spanner(label = "Living Standards", columns = d03_cp) |>
    gt::fmt_number(columns = c(d01_cp, d02_cp, d03_cp), decimals = .decimals) |>
    gt::tab_options(table.font.size = 12)
}

mpi_result$contribution$k_33 |>
  gt::gt() |>
  gt::tab_header(title = "Contribution to MPI by Indicator — 33% Poverty Cutoff") |>
  gtx()
```

### Headcount ratios

The **uncensored** headcount ratio shows the raw deprivation rate on each indicator, regardless of whether a household crosses the poverty line. The **censored** version counts only those households also identified as multidimensionally poor.

```{r, eval = FALSE}
mpi_result$headcount_ratio$uncensored   # deprivation rate — all households
mpi_result$headcount_ratio$k_33         # deprivation rate — poor households only
```

```{r, echo = FALSE}
mpi_result$headcount_ratio$uncensored |>
  dplyr::ungroup() |>
  gt::gt() |>
  gt::tab_header(title = "Uncensored Headcount Ratio (all households)") |>
  gtx(.decimals = 3)
```

```{r, echo = FALSE}
mpi_result$headcount_ratio$k_33 |>
  dplyr::ungroup() |>
  gt::gt() |>
  gt::tab_header(title = "Censored Headcount Ratio (poor households only, k = 33%)") |>
  gtx(.decimals = 3)
```

### The deprivation matrix

The deprivation matrix records each household's individual score and indicator flags. The first few rows:

```{r, eval = FALSE}
mpi_result$deprivation_matrix$uncensored |> head()
```

```{r, echo = FALSE}
mpi_result$deprivation_matrix$uncensored |>
  dplyr::ungroup() |>
  head() |>
  gt::gt() |>
  gt::tab_header(title = "Deprivation Matrix — first 6 households (uncensored)") |>
  gtx(.decimals = 0, .offset = 1) |>
  gt::fmt_number(columns = 3, decimals = 3)
```

After applying the poverty cutoff, households with a deprivation score below the threshold have all their indicator flags set to zero — they are not counted as poor:

```{r, eval = FALSE}
mpi_result$deprivation_matrix$k_33 |> head()
```

```{r, echo = FALSE}
mpi_result$deprivation_matrix$k_33 |>
  dplyr::ungroup() |>
  head() |>
  gt::gt() |>
  gt::tab_header(title = "Deprivation Matrix — first 6 households (k = 33% cutoff)") |>
  gtx(.decimals = 0, .offset = 1) |>
  gt::fmt_number(columns = 3, decimals = 3)
```

> To save memory, you can exclude deprivation matrices from the output by setting `include_deprivation_matrix = FALSE` in `compute_mpi()`.

---

## Step 5: Save results to Excel

`save_mpi()` writes all results to a formatted Excel workbook:

```{r, eval = FALSE}
save_mpi(mpi_result, mpi_specs = mpi_specs, filename = "MPI Results")
```

Each component gets its own sheet. To also include the specification table as a reference:

```{r, eval = FALSE}
save_mpi(
  mpi_result,
  mpi_specs = mpi_specs,
  filename = "MPI Results",
  include_specs = TRUE
)
```

---

## Quick-reference script

Here is the complete workflow in one place:

```{r, eval = FALSE}
library(mpindex)

# 1. Load specifications
mpi_specs <- global_mpi_specs(uid = "uuid", unit_of_analysis = 'households')

# 2. Compute the MPI
mpi_result <- compute_mpi(
  df_household,
  mpi_specs = mpi_specs,
  deprivations = list(
    nutrition = deprived(undernourished == 1 & age < 70, .data = df_household_roster, collapse_fn = max),
    child_mortality = deprived(with_child_died == 1),
    year_schooling = deprived(completed_6yrs_schooling == 2, .data = df_household_roster, collapse_fn = max),
    school_attendance = deprived(attending_school == 2 & age %in% 5:24, .data = df_household_roster, collapse_fn = max),
    cooking_fuel = deprived(cooking_fuel %in% c(4:6, 9)),
    sanitation = deprived(toilet > 1),
    drinking_water = deprived(drinking_water == 2),
    electricity = deprived(electricity == 2),
    housing = deprived(roof %in% c(5, 7, 9) |walls %in% c(5, 8, 9, 99) == 2 | floor %in% c(5, 6, 9)),
    assets = deprived(
      !(asset_tv + 
        asset_telephone + 
        asset_mobile_phone + 
        asset_computer + 
        asset_animal_cart + 
        asset_bicycle + 
        asset_motorcycle + 
        asset_refrigerator) > 1 & 
      (asset_car + asset_truck) > 0
    )
  )
)

# 3. Inspect results
mpi_result$index$k_33          # headline MPI, H, A
mpi_result$contribution$k_33   # indicator contributions (%)
mpi_result$headcount_ratio$k_33
mpi_result$deprivation_matrix$k_33

# 4. Save to Excel
save_mpi(mpi_result, mpi_specs = mpi_specs, filename = "MPI Results")
```

---

## Further reading

- [OPHI — How to Apply the Alkire-Foster Method](https://ophi.org.uk/research/multidimensional-poverty/how-to-apply-alkire-foster/)
- [Global MPI 2023 Report](https://hdr.undp.org/content/2023-global-multidimensional-poverty-index-mpi)
- Package documentation: `?compute_mpi`, `?define_mpi_specs`, `?define_deprivation`, `?save_mpi`
